SILVR: Guided Diffusion for Molecule Generation

### 2023-04-18 paper/manuscript.tex
Updates from Overleaf

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The abstract environment will automatically gobble the contents
%% if an abstract is not used by the target journal.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
Computationally generating novel synthetically accessible compounds with high affinity and low toxicity is a great challenge in drug-design. Machine-learning models beyond conventional pharmacophoric methods have shown promise in generating novel small molecule compounds, but require significant tuning for a specific protein target. Here, we introduce a method called selective iterative latent variable refinement (SILVR) for conditioning an existing a diffusion-based equivariant generative model without retraining. The model allows the generation of new molecules that fit into a binding site of a protein based on fragment hits. We use the SARS-CoV-2 Main Protease fragments from Diamond X-Chem that form part of the COVID Moonshot project as a reference dataset for conditioning the molecular generation. The SILVR rate controls the extent of conditioning and we show that moderate SILVR rates make it possible to generate new molecules of similar shape to the original fragments, meaning that the new molecules fit the binding site without knowledge of the protein. We can also merge up to 3 fragments into a new molecule without affecting the quality of molecules generated by the underlying generative model. Our method is generalizable to any protein target with known fragments and any diffusion based model for molecule generation. 
\end{abstract}

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=\textwidth]{paper/Figures/Fig1/Schematic_new.png}
    \caption{Schematic of the equivariant diffusion model with  selective iterative latent variable refinement (SILVR) indicated for every denoising step. Here the reference in blue on the left is 3 small fragments. They are evolved over time $t$ in the diffusion process to resemble a Gaussian distribution at $t=T$. The $\beta$ added represents the noise added at each step, and the multiple arrows in grey show the omitted steps from time $t=2$ to $t=T$. As atoms effectively 'diffuse' they can be perceived as changing position. To generate a new sample of a single molecule that links all fragments (left most yellow box), making use of the reference fragments, the EDM is sampled at each denoising step, as indicated by the EDM arrows, but now the reference at that same level of noise is added back in through SILVR at a given rate $r_{\mathbf{S}}$, until a new sample that resembles the reference is generated (following the bottom row along the yellow arrow).}
    \label{fig:schematic}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%
% Theory
%%%%%%%%%%%%%%%%%%%%%%%
\section{Theory}
\label{sec:theory}
\subsection*{Denoising diffusion probabilistic models (DDPM) as generative models}
DDPMs are often used as generative models that were developed for the generation of new images~\cite{sohl-dickstein2015deep, ho2020denoising, nichol2021improved}. More recently, the same idea has also been applied to molecular generation~\cite{hoogeboom2022equivariant}. The main idea behind diffusion models is for a neural network to learn the reverse of a diffusion process, often referred to as denoising, to \textit{sample} a new image or in our case a new molecule. In practice, this is done by training a neural network $\phi$ and generating samples $p(\mathbf{x}_{t-1}|\mathbf{x}_t)=\mathcal{N}(x_{t-1}; \mu_{\theta}(x_t;t), \sigma_t^2\mathbf{I})$ from  a Gaussian transition kernel with learned mean ($\mu_{\theta}(x_t;t)$) and variance ($\sigma_t$), with $\mathbf{x}_t$ being the data noised up to time $t$. Figure~\ref{fig:schematic} shows a schematic of the two main parts of such a DDPM, the diffusion part where noise is added for each timestep (shown in blue) and the denoising part (shown in yellow) that allows the generation of a new image (molecule).

The \textit{forward} part of the diffusion model is a Markov process: noise is added to a set of data $\mathbf{x}_0$ and over a time interval $t = [1,\ldots,T]$ according to the following distribution:
\begin{equation}
  q(\mathbf{x}_{1:T}|\mathbf{x}_0) = q(x_0)\prod_{t=1}^T q(\mathbf{x}_t|\mathbf{x}_{t-1}).
\end{equation}
The product of conditional probabilities $q(..)$ can be modelled as a Gaussian Kernel given by:
\begin{equation}
    q(\mathbf{x}_t, \mathbf{x}_s)=\mathcal{N}(x_t|\frac{\alpha_t}{\alpha_s}, \sigma_t^2-\frac{\alpha_t}{\alpha_s}\sigma^2_s),
\end{equation}
for any $t>s$. The parameters $\alpha_t \in \mathbb{R}^{+}$ specifies the amount of retained signal and $\sigma_t \in \mathbb{R}^{+}$ represents the variance and thus the amount of white noise added. The parameter $\alpha_t$ also directly relates to $\beta$ in Figure~\ref{fig:schematic} with $\alpha_t:=1-\beta_t$ and $\langle\alpha\rangle_t:=\prod_{s=1}^t\alpha_s$. $\beta_t$ is a fixed variance schedule which adds noise with each timestep $t$.
Different researchers have examined different noise schedules~\cite{sohl-dickstein2015deep, ho2020denoising}.
% Next talk about the denoising step
$p_{\theta}$ is the parametric probability distribution we want to learn. This generative process is given by:

\begin{equation}
p_{\theta}(\mathbf{x}_{t-1}|\mathbf{x}_t)=\mathcal{N}(x_{t-1}; \mu_{\theta}(x_t;t), \sigma_t^2\mathbf{I}).
\end{equation}
A sample for denoising timestep $t-1$ can be generated from the following equation given the neural network $\phi$ that has been trained on the diffusion process: 
\begin{equation}
 x_{t-1} = \frac{1}{\alpha_t}(x_t-\frac{1-\alpha_t}{\sqrt{1-\langle\alpha_t\rangle}}\phi(x_t,t)) +\sigma \mathbf{z},   
\end{equation}

with $\mathbf{z}\sim\mathcal{N}(0,\mathbf{I})$.

The process is iterated until $t=1$ and as such a new sample $x_0$ of the denoising process---which is intended to represent a proposed molecule design---is generated.

%%%%%%%
% EDMs
%%%%%%%
\subsection*{Equivariant diffusion model (EDM)}
In the previous section introduced the neural network $\phi(\mthabf{x},t)$. In practice, it makes sense to use an euqivariant graph neural network, as it is a data efficient way to learn about molecules. If a model has rotational and translational equivarience, it means a neural network does not need to learn orientations and transations of molecules from scratch. We chose the EDM by Hoogeboom \textit{et al.}~\cite{hoogeboom2022equivariant} as our baseline generative model, as it provides a generative model for new molecules and has equivariance already built into it. Furthermore, it has all code and pretrained weights available online~\cite{githubhoogeboom}. The basic concept behind equivarience is that the model is invariant to rotations and translation, in this case the E(3) group. This means that scalar (features such as atom types) and vector node properties (such as the positions) will be invariant to group transformations. As a result the order in which a rotation is applied does not matter. The input to the model can be rotated and diffusion/denoising  applied to get a structure, or the diffusion/denoising process can be applied first followed by the same rotation to get the same output. 
Mathematically this means that if we have a set of point $\mathbf{x} = (\mathbf{x}_1,\ldots,\mathbf{x}_N) \in \mathbb{R}^{N\times 3}$ and each of these points has an associated set of scalar feature vectors $\mathbf{h}\in \mathbb{R}^{N\times k}$ which are invariant to group transformations. The position translations and rotations are defined according to the orthogonal matrix: $\mathbf{Rx + t} = (\mathbf{R_{x_1}+t},\ldots, \mathbf{R_{x_N}+t})$. Satorras \textit{et al.}~\cite{satorras2022equivariant} have proposed an E(n) Equivariant Graph Neural Network on which the EDM by Hoogeboom \textit{et al.} builds. For full details on the architecture and their code see~\cite{hoogeboom2022equivariant}. 
%Something on latent variables Z.

%%%%
% Subsection: ILVR
%%%%
\subsection{Iterative Latent Variable Refinement as a conditioning tool}
Choi \textit{et al.}~\cite{choi2021ilvr} introduced a way to condition DDPMs by sampling their images from the conditional distribution $p(x_0|c)$, with the condition c:
\begin{equation}
    P_\theta(x_0|c) = \int p_\theta(x_{0:T}|c)Dx_{1:T}
\end{equation}
\begin{equation}
    p_{\theta}(x_{0:T}|c)=p(x_T)\prod_{t=1}^{T}p_{\theta}(x_{t-1}x_t,c)
\end{equation}
This iterative latent variable refinement (ILVR) means that the condition can be applied during the denoising steps without additional training. In the case of Choi et al, they used downsampled reference images as conditioning in order to generate novel images similar to the reference.

%%%%
% Subsection: SILVR
%%%%
\subsection{Selective Iterative Latent Variable Refinement (SILVR)}
\label{sec:silvr}
We propose a new method, that combines ILVR by Choi \textit{et al.}~\cite{choi2021ilvr} proposed in an image generation context and the EDM by Hoogeboom \textit{et al.}. This allows us to generate a selective iterative latent variable refinement (SILVR) procedure in which we introduce information of a reference molecule into the denoising process. We describe this method as "selective" due to the ability to guide individual atoms at independent rates. The reference can be a single molecule or a series of superimposed fragments, and additional unguided dummy atoms can also be defined at the beginning of the denoising process. As a result, we propose the following algorithm for SILVR~\ref{alg:silvr}.  We consider latent space variables $\mathbf{z}=[\mathbf{z}^{(x)},\mathbf{z}^{(h)}]$ for the standard denoising process and  $\mathbf{\hat{z}} = [\mathbf{\hat{z}}^{(y)},\mathbf{\hat{z}}^{(h)}]$, for the set of reference coordinates given by $\mathbf{y}$. The vector $\mathbf{h}$ contains all the scalar node properties of the equivariant EDM. 
\begin{algorithm}
\caption{SILVR}\label{alg:silvr}
\begin{algorithmic}[1]
\State \textbf{Input:} Reference molecule $\mathbf{y,h}$, EDM $\kappa$ 
\State \textbf{Output:} Generated molecule $\mathbf{x,h}$

\State Compute $\mathbf{\hat{z}}_0$ from $\mathbf{y, h}$, such that $[\mathbf{z}^{(y)} , \mathbf{z}^{(h)} ] = f (\mathbf{y, h})$ is E(3). 
\State Subtract center of geometry (COG) from $\mathbf{\hat{z}}_0$
\Comment{Center reference at zero}
\State Sample $z_T \sim \mathcal{N}(\mathbf{0,I})$
\For {$t$ = $T,...,1$}
    \State Sample $\boldsymbol{\epsilon} \sim \mathcal{N}(\mathbf{0,I})$
    \State Subtract COG from $\boldsymbol{\epsilon}^{(x)}$
    \State $\mathbf{\hat{z}}_{t-1} = \alpha_{t-1} \mathbf{\hat{z}}_{0} +\sigma_{t-1}\times\boldsymbol{\epsilon}$
    \Comment{Noise reference to t=t-1}
    \State $\mathbf{z}_{t-1} \gets \kappa (\mathbf{z}_t,t)$
    \Comment{Compute denoising step}
    \State Update $\mathbf{z}_{t-1} \gets \mathbf{z}_{t-1} - \alpha_{t-1}r_{\mathrm{S}} \mathbf{z}_{t-1} +r_{\mathrm{S}} \mathbf{\hat{z}}_{t-1}$ \Comment{SILVR equation}
    \State Subtract COG from updated $\mathbf{z}_{t-1}$

@@

The core of the new method is the addition of a refinement step within the denoising process during runtime of any pre-trained E(3) EDM. The resulting SILVR model produces conditional samples without any conditional training when generating new molecules. Figure~\ref{fig:silvr_explanation} shows an illustrative example of how the SILVR rate $r_{\mathrm{S}}$ is used to shift the latent space vector $\mathbf{z}_{t-1}$ at any point in the denoising process from $T\ldots t=1$. Using the SILVR equation as defined in the SILVR algorithm~\ref{alg:silvr} line 11, an existing denoising step (green) is brought closer to the reference (purple) in latent space.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{paper/Figures/Fig2/fig2.jpg}
    \caption{Schematic illustrating the influence of applying the SILVR equation~\ref{eq:SILVR} to molecules in the latent space. At each denoising step if $r_{\mathrm{S}}>0$ the latent space vector gets scaled by the SILVR rate. This is done from the first denoising timestep (top) until the last one (bottom). Repeatedly applying SILVR will result in a molecule that resembles the references for $r_{\mathrm{S}}>0$ (right) and does not for $r_{\mathrm{S}}=0$ (left)}
    \label{fig:silvr_explanation}
\end{figure}

@@

\subsection{Reference Dataset: COVID Moonshot}
Reference molecules were selected from the original 23 non-covalent hits of the SARS-CoV-2 Main Protease (Mpro) identified by the XChem fragment screen~\cite{douangamath2020crystallographic} as part of the Covid Moonshot Project~\cite{consortium2023open, consortium2021open}. A more detailed picture of all fragments is presented in Figure~\ref{fig:moonshot_data} of the Supplementary information(SI). Fragments were visualised using NGLview version 3.03 and combinations were arbitrarily selected as test cases for the following different experimental settings for trying to understand the similarity between the reference and the new sample. We looked at the following scenarios:
\begin{enumerate}
    \item Using three distinct fragments with substantial overlap as a reference to generate new samples
    \item Using two fragments with a slight overlap to generate a single new sample.
    \item And using two fragments that are disconnected to investigate linker generation
\end{enumerate}
Fragments \texttt{x0072} and \texttt{x0354} were selected for benchmarking the effect of the SILVR rate $r_{\mathrm{S}}$ on sampling; \texttt{x1093}, \texttt{x0072}, \texttt{x2193} were randomly chosen to represent three significantly overlapping fragments; \texttt{x0434}, \texttt{x0305} and \texttt{x0072}, \texttt{x0354} were used as partially overlapping fragments; and \texttt{x0874}, \texttt{x0397} were used as two disconnected fragments, resembling a linker design type problem. The bonding information of selected fragments was deleted and cartesian coordinates were combined into a single \texttt{XYZ} file. Values of $r_{\mathrm{S}}$ were selected and added to the \texttt{XYZ} file to create a reference file containing all experiment setup information. Each experiment was sampled 1000 times.

@@

\subsubsection{Shape complementarity of sample to protein}
The agreement in the shape of samples and the main protease binding site was determined using openeye-toolkits version 2022.2.2 Gaussian scoring function \textit{shapegauss}~\cite{openeyescientificsoftwareinc.oedocking,kelley2015posit}. This scoring function measures the shape complementarity between the ligand and receptor by considering each heavy atom as a Gaussian function~\cite{mcgann2003gaussian}. The most favourable score occurs “when two atoms touch but do not overlap”.  This metric does not consider any intermolecular interactions beyond shape complementarity. The protein receptor file was prepared from the Mpro-x0072 crystal structure with removal of the ligand. The \texttt{XYZ} coordinates of samples were directly read into openeye-toolkit and the pose was re-scored with \textit{shapegauss}.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The abstract environment will automatically gobble the contents
%% if an abstract is not used by the target journal.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
Computationally generating novel synthetically accessible compounds with high affinity and low toxicity is a great challenge in drug-design. Machine-learning models beyond conventional pharmacophoric methods have shown promise in generating novel small molecule compounds, but require significant tuning for a specific protein target. Here, we introduce a method called selective iterative latent variable refinement (SILVR) for conditioning an existing a diffusion-based equivariant generative model without retraining. The model allows the generation of new molecules that fit into a binding site of a protein based on fragment hits. We use the SARS-CoV-2 Main protease fragments from Diamond X-Chem that form part of the COVID Moonshot project as a reference dataset for conditioning the molecular generation. The SILVR rate controls the extent of conditioning and we show that moderate SILVR rates make it possible to generate new molecules of similar shape to the original fragments, meaning that the new molecules fit the binding site without knowledge of the protein. We can also merge up to 3 fragments into a new molecule without affecting the quality of molecules generated by the underlying generative model. Our method is generalizable to any protein target with known fragments and any diffusion based model for molecule generation. 
\end{abstract}

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=\textwidth]{paper/Figures/Fig1/Schematic_new.png}
    \caption{Schematic of the equivariant diffusion model with  selective iterative latent variable refinement (SILVR) indicated for every denoising step. Here the reference in blue on the left is 3 small fragments. They are evolved over time $t$ in the diffusion process to resemble a Gaussian distribution at $t=T$, see Equation~\ref{eqn:noising}. The $\beta$ added represents the noise added at each step, and the dots show the omitted steps from time $t=3$ to $t=T$. As atoms effectively 'diffuse' they can be perceived as changing position. To generate a new sample a sample is generated from $p_{\theta}(x)$ according to Equation~\ref{eqn:denoising}, this distribution is from the learned EDM. At each denoising step, a set of reference fragments ($\mathbf{y}_t$) at that same level of noise $t$ is used which is indicated by the SILVR arrows to condition the EDM. This is controlled through SILVR at a given rate $r_{\mathbf{S}}$, until a new sample that resembles the reference is generated (following the bottom row along the yellow boxes and EDM arrows).}
    \label{fig:schematic}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%
% Theory
%%%%%%%%%%%%%%%%%%%%%%%
\section{Theory}
\label{sec:theory}
\subsection*{Denoising diffusion probabilistic models (DDPM) as generative models}
DDPMs are often used as generative models that were developed for the generation of new images~\cite{sohl-dickstein2015deep, ho2020denoising, nichol2021improved}. More recently, the same idea has also been applied to molecular generation~\cite{hoogeboom2022equivariant}. The main idea behind diffusion models is for a neural network to learn the reverse of a diffusion process, often referred to as denoising, to \textit{sample} a new image or in our case a new molecule. In practice, this is done by training a neural network $\phi$ and generating samples $p(\mathbf{x}_{t-1}|\mathbf{x}_t)=\mathcal{N}(x_{t-1}; \mu_{\theta}(x_t;t), \sigma_t^2\mathbf{I})$ from  a Gaussian transition kernel with learned mean ($\mu_{\theta}(x_t;t)$) and variance ($\sigma_t$), with $\mathbf{x}_t$ being the data noised up to time $t$. Figure~\ref{fig:schematic} shows a schematic of the two main parts of such a DDPM, the diffusion part, where noise is added for each timestep (shown in blue) and the denoising part (shown in yellow) that allows the generation of a new image (molecule).

The \textit{forward} part of the diffusion model is a Markov process: noise is added to a set of data $\mathbf{x}_0$ and over a time interval $t = [1,\ldots,T]$ according to the following distribution:
\begin{equation}
  q(\mathbf{x}_{1:T}|\mathbf{x}_0) = q(x_0)\prod_{t=1}^T q(\mathbf{x}_t|\mathbf{x}_{t-1}).
  \label{eqn:noising}
\end{equation}
The product of conditional probabilities $q(\mathbf{x}_t | \mathbf{x}_{t-1})$ can be modelled as a Gaussian Kernel given by:
\begin{equation}
q(\mathbf{x}_t | \mathbf{x}_{t-1}) = \mathcal{N}(\mathbf{x}_t;\sqrt{1-\beta_t}\mathbf{x}_{t-1}, \beta_t\mathbf{I}),
\end{equation}
where the mean of the normal distribution is given by $ \sqrt{1-\beta_t}$ and $\beta_t$ a fixed variance. 
In order to diffuse directly to timestep $s$ in the diffusion process the following shorthand is possible:
\begin{equation}
    q(\mathbf{x}_t, \mathbf{x}_s)=\prod_{t=1}^s q(\mathbf{x}_t|\mathbf{x}_{t-1})=\mathcal{N}(\mathbf{x}_t|\frac{\alpha_t}{\alpha_s}, \sigma_t^2-\frac{\alpha_t}{\alpha_s}\sigma^2_s),
\end{equation}
for any $t>s$. The parameters $\alpha_t \in \mathbb{R}^{+}$ specifies the amount of retained signal and $\sigma_t \in \mathbb{R}^{+}$ represents the variance and thus the amount of white noise added. The parameter $\alpha_t$ also directly relates to $\beta$ in Figure~\ref{fig:schematic} with $\alpha_t:=1-\beta_t$ and $\langle\alpha\rangle_t:=\prod_{s=1}^t\alpha_s$. $\beta_t$ is a fixed variance schedule which adds noise with each timestep $t$.
Different researchers have examined different noise schedules~\cite{sohl-dickstein2015deep, ho2020denoising}.
% Next talk about the denoising step
What we are actually interested in is learning the reverse process, i.e. the denoising and generating $\mathbf{x}$, however the reverse of the process $q(\mathbf{x}_{t-1}|\mathbf{x}_{t=T})$ is intractable. The DDPM learns this reverse transitions $p_{\theta}(\hat{\mathbf{x}}_{t-1}|\mathbf{x}_{t=T})$, which is also a Gaussian transition kernel. This generative (or denoising) process is given by:

\begin{equation}
p_{\theta}(\hat{\mathbf{x}}_{t-1}|\mathbf{x}_{t=T})=\mathcal{N}(\hat{\mathbf{x}}_{t-1}; \mu_{\theta}(\mathbf{x}_t;T), \sigma_t^2\mathbf{I}).
\label{eqn:denoising}
\end{equation}
$\mu_{\theta}$, is the learned mean and $\sigma$ represents a fixed variance for this transition process.
A sample for denoising timestep $t-1$ can be generated from the following equation given the neural network $\phi(\mathbf{x}_t,t)$ that has been trained on the diffusion process: 
\begin{equation}
 \hat{\mathbf{x}}_{t-1} = \frac{1}{\alpha_t}(\mathbf{x}_t-\frac{1-\alpha_t}{\sqrt{1-\langle\alpha_t\rangle}}\phi(\mathbf{x}_t,t)) +\sigma \rho,   
\end{equation}

with $\rho\sim\mathcal{N}(0,\mathbf{I})$.

The process is iterated until $t=1$ and as such a new sample $\mathbf{x}_0$ of the denoising process---which is intended to represent a proposed molecule design---is generated.

%%%%%%%
% EDMs
%%%%%%%
\subsection*{Equivariant diffusion model (EDM)}
In the previous section introduced the neural network $\phi(\mathbf{x}_t,t)$. In practice, it makes sense to use an euqivariant graph neural network, as it is a data efficient way to learn about molecules. If a model has rotational and translational equivarience, it means a neural network does not need to learn orientations and transations of molecules from scratch. We chose the EDM by Hoogeboom \textit{et al.}~\cite{hoogeboom2022equivariant} as our baseline generative model, as it provides a generative model for new molecules and has equivariance already built into it. Furthermore, it has all code and pretrained weights available online~\cite{githubhoogeboom}. The basic concept behind equivarience is that the model is invariant to rotations and translation, in this case the E(3) group. This means that scalar (features such as atom types) and vector node properties (such as the positions) will be invariant to group transformations. As a result the order in which a rotation is applied does not matter. The input to the model can be rotated and diffusion/denoising  applied to get a structure, or the diffusion/denoising process can be applied first followed by the same rotation to get the same output. 
Mathematically this means that if we have a set of point $\mathbf{x} = (\mathbf{x}_1,\ldots,\mathbf{x}_N) \in \mathbb{R}^{N\times 3}$ and each of these points has an associated set of scalar feature vectors $\mathbf{h}\in \mathbb{R}^{N\times k}$ which are invariant to group transformations. The position translations and rotations are defined according to the orthogonal matrix: $\mathbf{Rx + t} = (\mathbf{R_{x_1}+t},\ldots, \mathbf{R_{x_N}+t})$. Satorras \textit{et al.}~\cite{satorras2022equivariant} have proposed an E(n) Equivariant Graph Neural Network on which the EDM by Hoogeboom \textit{et al.} builds. 
The actual diffusion process of the Hoogeboom model relies taking place on a set of points $M$ $\{\{\mathbf{x}_i,\mathbf{h}_i\}\}_{i\ldots M}$ on a latent variable that combines atom coordinates $\mathbf{x}_i$ and node features of each atom $\mathbf{h}_i$, such that: $\mathbf{z}_t = [\mathbf{z}_t^{(x), \mathbf{z}_t^{(h)}}$. The diffusion process similar to that of equation~\ref{eqn:noising} can be written as:
\begin{equation}
    q(\mathbf{z}_t| \mathbf{x},\mathbf{h}) = \mathcal{N}_{xh}(\mathbf{z}_t|\alpha_t[\mathbf{x,h}],\sigma^2_t\mathbf{I}].
\end{equation}
Similarly for the genreative denoising process the distribution can be written as:
\begin{equation}
    p_\theta(\mathbf{z}_s| \mathbf{z}_t) = \mathcal{N}_{xh}(\mathbf{z}_s | \mu_{\theta t\rightarrow s}( [ \mathbf{\hat{x},\hat{h}}],\mathbf{z}_t),\sigma_{t\rightarrow s}^2\mathbf{I}).
\end{equation}
This is the equivlanet of equation~\ref{eqn:denoising}, using $\mathbf{\hat{x}}, \mathbf{\hat{h}}$ as the data variables that are estimated by the neural network. The neural network $\phi(\mathbf{z}_t,t)$ outputs an auxillary variable $\hat\epsilon = [\hat{\epsilon}^{(x)},\hat{\epsilon}^{(h}]$, from which $\mathbf{\hat{x}}, \mathbf{\hat{h}}$ can be recovered as:
\begin{equation}
    [\mathbf{\hat{x}}, \mathbf{\hat{h}}] = \frac{\mathbf{z}_t}{\alpha_t}-\hat{\epsilon}_t\frac{\sigma_t}{\alpha_t}.
\end{equation}
We use the notation $\kappa$ to generate a sample [\mathbf{\hat{x}}, \mathbf{\hat{h}}] from the EDM. 
For full details on the architecture and their code see~\cite{hoogeboom2022equivariant}. 
%Something on latent variables Z.

%%%%
% Subsection: ILVR
%%%%
\subsection{Iterative Latent Variable Refinement as a conditioning tool}
Conditioning diffusion models is often desirable to for example generate similar images to that of an original input images, or in the example of Hoogeboom generating molecules in the presence of an external electric field resulting in more polarizable molecules. However, this conditioning requires retraining the network to accommodate for this condition. Choi \textit{et al.}~\cite{choi2021ilvr} introduced a way to condition DDPMs without having to retrain the neural network. In the generative process it is possible to introduce a condition $c$ using a conditional distribution $p(x_0|c)$:
\begin{equation}
    p_\theta(x_0|c) = \int p_\theta(x_{0:T}|c)dx_{1:T}
\end{equation}
\begin{equation}
    p_{\theta}[x_{\{0:T\}}|c]=p(x_T)\prod_{t=1}^{T}p_{\theta}(x_{t-1}|x_t,c)
\end{equation}
Their trick is to use a reference image $\mathbf{y}$ and place it in the same downsampled fillter $\psi(\mathbf{y})$ as the generated image $\psi(\mathbf{\hat{x}}_0)$. This means the original image and the reference are  in the same latent space. So in each denoising step the proposal distribution is matched with that of the reference $y$ noised up to the appropriate timestep. An unconditional distribution at $t$ is generated first:
\begin{equation}
\mathbf{\hat{x}}_{t-1} \sim p_{\theta}( \mathbf{\hat{x}}|\mathbf{x}_t),
\end{equation}
then this  new sample is 'adjusted' according to:
\begin{equation}
    \mathbf{\hat{x}}^{\prime}_{t-1} = \psi(\mathbf{y}_{t-1}) + (I-\psi)(\mathbf{\hat{x}}_{t-1})
\end{equation}
This iterative latent variable refinement (ILVR) means that the condition can be applied during the denoising steps without additional training. In the case of Choi \textit{et al.}~\cite{choi2021ilvr}, they used downsampled reference images as conditioning in order to generate novel images similar to the reference.

%%%%
% Subsection: SILVR
%%%%
\subsection{Selective Iterative Latent Variable Refinement (SILVR)}
\label{sec:silvr}
We propose a new method, that combines ideas from ILVR by Choi \textit{et al.}~\cite{choi2021ilvr} proposed in an image generation context and the EDM by Hoogeboom \textit{et al.}~\cite{hoogeboom2022equivariant}. This allows us to generate a selective iterative latent variable refinement (SILVR) procedure in which we introduce information of a reference molecule into the denoising process. We describe this method as "selective" due to the ability to guide individual atoms at independent rates. The reference can be a single molecule or a series of superimposed fragments, and additional unguided dummy atoms can also be defined at the beginning of the denoising process. We consider latent space variables $\mathbf{z}=[\mathbf{z}^{(x)},\mathbf{z}^{(h)}]$ for the standard denoising process and  $\mathbf{\tilde{z}} = [\mathbf{z}^{(y)},\mathbf{z}^{(h)}]$, for the set of reference coordinates given by $\mathbf{y}$. The vector $\mathbf{h}$ contains all the scalar node properties of the equivariant EDM for the reference. 
Similarly to Choi \texit{et al.}, we update the diffusion process at noise level $t$ in the latent space $\mathbf{z}$, with the reference $\mathbf{\tilde{z}}$ using a factor, or vector if used at variable rates for different atoms $r_{\mathrm{S}}$. We call $r_{\mathrm{S}}$ the SILVR rate and this leads to an overall update or conditioning towards a reference at each step in the generative denoising process according to the SILVR equation~\ref{eqn:SILVR}.
\begin{equation}
\mathbf{z}_{t-1}^{\prime} = \mathbf{z}_{t-1}-\alpha_{t-1}r_{\mathrm{S}}\mathbf{z}_{t-1}+r_{\mathrm{S}}\mathbf{\tilde{z}}_{t-1}
    \label{eqn:SILVR}
\end{equation}
As a result, we propose the following algorithm for the generation of condition samples according to SILVR~\ref{alg:silvr}.
\begin{algorithm}
\caption{SILVR}\label{alg:silvr}
\begin{algorithmic}[1]
\State \textbf{Input:} Reference molecule $\mathbf{y,h}$, EDM $\kappa$ 
\State \textbf{Output:} Generated molecule $\mathbf{x,h}$

\State Compute $\mathbf{\tilde{z}}_0$ from $\mathbf{y, h}$, such that $[\mathbf{z}^{(y)} , \mathbf{z}^{(h)} ] = f (\mathbf{y, h})$ is E(3). 
\State Subtract center of geometry (COG) from $\mathbf{\tilde{z}}_0$
\Comment{Center reference at zero}
\State Sample $z_T \sim \mathcal{N}(\mathbf{0,I})$
\For {$t$ = $T,...,1$}
    \State Sample $\boldsymbol{\epsilon} \sim \mathcal{N}(\mathbf{0,I})$
    \State Subtract COG from $\boldsymbol{\epsilon}^{(x)}$
    \State $\mathbf{\hat{z}}_{t-1} = \alpha_{t-1} \mathbf{\tilde{z}}_{0} +\sigma_{t-1}\times\boldsymbol{\epsilon}$
    \Comment{Noise reference to t=t-1}
    \State $\mathbf{z}_{t-1} \gets \kappa (\mathbf{z}_t,t)$
    \Comment{Compute denoising step}
    \State Update $\mathbf{z}^{\prime}_{t-1} \gets \mathbf{z}_{t-1} - \alpha_{t-1}r_{\mathrm{S}} \mathbf{z}_{t-1} +r_{\mathrm{S}} \mathbf{\tilde{z}}_{t-1}$ \Comment{SILVR equation}
    \State Subtract COG from updated $\mathbf{z}_{t-1}$

@@

The core of the new method is the addition of a refinement step within the denoising process during runtime of any pre-trained E(3) EDM. The resulting SILVR model produces conditional samples without any conditional training when generating new molecules. Figure~\ref{fig:silvr_explanation} shows an illustrative example of how the SILVR rate $r_{\mathrm{S}}$ is used to shift the latent space vector $\mathbf{z}_{t-1}$ at any point in the denoising process from $T\ldots t=1$, here in 2D for illustration purposes. Using the SILVR equation~\ref{eqn:SILVR}, an existing denoising step (light blue) is brought closer to the reference (purple) in the latent space according to a scaled version of purple using $r_{\mathrm{S}}$ (green).

@@

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{paper/Figures/Fig2/fig2.jpg}
    \caption{Schematic illustrating the influence of applying the SILVR equation~\ref{eqn:SILVR} to molecules in the latent space. At each denoising step if $r_{\mathrm{S}}>0$ the latent space vector gets scaled by the SILVR rate. This is done from the first denoising timestep (top) until the last one (bottom). Repeatedly applying SILVR will result in a molecule that resembles the references for $r_{\mathrm{S}}>0$ (right) and does not for $r_{\mathrm{S}}=0$ (left)}
    \label{fig:silvr_explanation}
\end{figure}

@@

\subsection{Reference Dataset: COVID Moonshot}
Reference molecules were selected from the original 23 non-covalent hits of the SARS-CoV-2 Main Protease (Mpro) identified by the XChem fragment screen~\cite{douangamath2020crystallographic} as part of the COVID Moonshot Project~\cite{consortium2023open, consortium2021open}. A more detailed picture of all fragments is presented in Figure~\ref{fig:moonshot_data} of the Supplementary information(SI). Fragments were visualised using NGLview version 3.03 and combinations were arbitrarily selected as test cases for the following different experimental settings for trying to understand the similarity between the reference and the new sample. We looked at the following scenarios:
\begin{enumerate}
    \item Using three distinct fragments with substantial overlap as a reference to generate new samples.
    \item Using two fragments with a slight overlap to generate a single new sample.
    \item And using two fragments that are disconnected to investigate linker generation.
\end{enumerate}
Fragments \texttt{x0072} and \texttt{x0354} were selected for benchmarking the effect of the SILVR rate $r_{\mathrm{S}}$ on sampling; \texttt{x1093}, \texttt{x0072}, \texttt{x2193} were randomly chosen to represent three significantly overlapping fragments; \texttt{x0434}, \texttt{x0305} and \texttt{x0072}, \texttt{x0354} were used as partially overlapping fragments; and \texttt{x0874}, \texttt{x0397} were used as two disconnected fragments, resembling a linker design type problem. The bonding information of selected fragments was deleted and cartesian coordinates were combined into a single \texttt{XYZ} file. Values of $r_{\mathrm{S}}$ were selected and added to the \texttt{XYZ} file to create a reference file containing all experiment setup information. Each experiment was sampled 1000 times.

@@

\subsubsection{Shape complementarity of sample to protein}
The agreement in the shape of samples and the MPro binding site was determined using openeye-toolkit version 2022.2.2 Gaussian scoring function \textit{shapegauss}~\cite{openeyescientificsoftwareinc.oedocking,kelley2015posit}. This scoring function measures the shape complementarity between the ligand and receptor by considering each heavy atom as a Gaussian function~\cite{mcgann2003gaussian}. The most favourable score occurs “when two atoms touch but do not overlap”.  This metric does not consider any intermolecular interactions beyond shape complementarity. The protein receptor file was prepared from the Mpro-x0072 crystal structure with removal of the ligand. The \texttt{XYZ} coordinates of samples were directly read into the openeye-toolkit and the pose was re-scored with \textit{shapegauss}.

\section{Results}
In the following, we will demonstrate how SILVR can be used to generate conditioned samples to a reference using a pre-trained EDM without additional training. The main questions we set out to answer with SILVR were:
\begin{enumerate}
    \item Can we generate samples from the EDM that are similar to the reference structures?
    \item Is there a SILVR rate $r_{\mathrm{S}}$ that will provide enough diversity while still retaining reference features?
    \item Do the generated samples of new molecules still fit into the Mpro binding site?
    \item Can we link molecule fragments without incorporating binding site information as part of additional training?
\end{enumerate}
